---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

The assignment instructions say to always use echo = TRUE. This is the default, so no need to set it or alter it. 

## Loading and preprocessing the data

```{r}
#Source data .zip file (repdata_data_activity.zip) downloaded from link 
#on Coursera website: https://d396qusza40orc.cloudfront.net
#Assumption: .zip has been unzipped and the .csv it contains is located in 
#the same directory as the .Rmd file.
df <- read.csv("activity.csv")
```

## What is mean total number of steps taken per day?
First, make sure the dplyr package is loaded. 
```{r}
require(dplyr)
```
Now create a new dataframe with total number of steps per day.

```{r}
df2 <- df %>% group_by(date) %>% summarize(Total_Steps = sum(steps))
```

Create a histogram. Since I realized I do not understand the difference 
between a histogram and a barplot (doh!), I googled around and read this:
http://www.forbes.com/sites/naomirobbins/2012/01/04/a-histogram-is-not-a-bar-chart/#f6aefc328aff, which propelled me to create my histogram as follows.

```{r}
hist(df2$Total_Steps,xlab="Total Steps Per Day",ylab="Days", ylim = range(0,40),main="Total Steps Histogram, Subject X, October-November 2012",col = "blue")
```

Next, find and report the mean and median of the total number of steps per day. 
Note that we need to ignore NA records, of which there are 8 in the dataset. 
Mean:

```{r}
mean(df2$Total_Steps,na.rm = T)
```
And median:
```{r}
median(df2$Total_Steps,na.rm = T)
```

## What is the average daily activity pattern?

Plot the relationship of 5-minute intervals to average number of steps. Need to filter out the NAs. 

```{r}
df3 <- subset(df,!is.na(steps)) %>% group_by(interval) %>% summarize(Average_Steps = mean(steps))
with(df3,plot(interval,Average_Steps,type="l",xlab="5-minute interval", ylab="Average Steps"))
```

Next, we want to identify which interval has the maximum Average_Steps value. 

```{r}
interval_with_max_steps <- filter(df3, Average_Steps == max(Average_Steps))
interval_with_max_steps$interval
```

## Imputing missing values

Let's find the number of rows that contain missing values (NAs).

```{r}
df_with_na <- df[!complete.cases(df),]
str(df_with_na)
```

OK, now update the original data.frame where the steps column is "NA" with the average for the corresponding interval value. This seems a bit dubious, but I'm doing what I'm told to do. 

Reminder that "df" is the untouched R data.frame representation of activity.csv, and
"df3" is the average number of steps by interval (with NAs ignored).

```{r}
#Merge the two DFs on interval. This will give us a fourth Average_Steps column.
df4 <- merge(df,df3,by="interval")
#Update the steps column when NA with the average for that interval.
df4$steps <- ifelse(is.na(df4$steps),df4$Average_Steps,df4$steps)
```
Present the same histogram and mean/median values as above, only this time after having updated all the NAs in the step column. 

```{r}
df5 <- df4 %>% group_by(date) %>% summarize(Total_Steps = sum(steps))
```
The revised histogram:

```{r}
hist(df5$Total_Steps,xlab="Total Steps Per Day",ylab="Days", ylim = range(0,40),main="Total Steps Histogram (updated NAs), Subject X, October-November 2012",col = "blue")
```

And the revised mean and median:

```{r}
mean(df5$Total_Steps)
median(df5$Total_Steps)
```

## Are there differences in activity patterns between weekdays and weekends?
